Maximum Entropy Prediction Markets

### 2014-02-11 WriteUp/liquidity.tex
some edits to liquidity
fixed lb proof in liquidity write-up

% Metadata Information
\acmVolume{9}
\acmNumber{4}
\acmArticle{39}
\acmYear{2013}
\acmMonth{6}
\usepackage[numbers]{natbib}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{amsthm}
\usepackage{parskip}
\usepackage{booktabs}
%new commands
\newcommand{\la}{\langle}
\newcommand{\ra}{\rangle}
\newcommand{\mX}{\mathcal{X}}
\newcommand{\bR}{\mathbf{R}}
\newcommand{\grad}{\nabla}
\newcommand{\Exp}{\mathbf{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\q}{\mathbf{q}}
\newcommand{\mM}{\ensuremath{\mathcal{M}}}
\newcommand{\hmu}{\ensuremath{\hat{\mu}}}
\newcommand{\mP}{\ensuremath{\mathcal{P}}}
\newcommand{\mY}{\ensuremath{\mathcal{Y}}}
\newcommand{\mS}{\ensuremath{\mathcal{S}}}
\newcommand{\ds}{\displaystyle}
\newcommand{\mF}{\mathcal{F}}
\newcommand{\mD}{\mathcal{D}}
\newcommand{\htheta}{\hat{\theta}}
%---------------------------------------------------------------------------
\newtheorem{thm}{Theorem}[section]
\newtheorem{prop}[thm]{Proposition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{con}[thm]{Conjecture}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{ex}[thm]{Example}
\newtheorem{qn}[thm]{Question}
\newcommand{\defn}{\bigbreak\noindent{\bf Definition. }}
\newcommand{\undefn}{\bigbreak}
\newcommand{\rmk}{\bigbreak\noindent{\bf Remark. }}
\newcommand{\unrmk}{\bigbreak}
%\newcommand{\proof}{\noindent{\bf Proof. }}
\newcommand{\unproof}{\hfill$\Box$\bigbreak}
\newcommand{\defeq}{\stackrel{\mathrm{def}}{=}}
\newcommand{\holds}[1]{\stackrel{?}{#1}}
\newcommand{\conv}{\mathrm{conv}}
\newcommand{\ie} {{\it i.e. }}
\newcommand{\eg} {{\it e.g. }}
\newcommand{\etal} {{\it et al. }}
\newcommand{\E}{\mathbf{E}}
\newcommand{\betavec}{\pmb{\beta}}
\newcommand{\qvec}{\mathbf{q}}
\newcommand{\lpf}{\mathbf{\psi}} %logpartition function
\newcommand{\family}{\mathcal F} 
\newcommand{\suff}{\pmb{\phi}} % suff stats
\newcommand{\muvec}{\pmb{\mu}}

@@

Thus, the expected payoff  for a trader with exponential family belief parameter $\beta$ who moves the market state from $\theta$ to $\theta'$ can be written as
\begin{eqnarray*}
\E_{x\sim p(x;\beta)}\left[C_{\lambda}(\theta)-\theta\phi(x)-(C_{\lambda}(\theta')-\theta'\phi(x))\right]=\lambda[D_{T}(\theta/\lambda,\beta)-D_{T}(\theta'/\lambda,\beta)]
\end{eqnarray*}¥
If $\theta'=a\beta+(1-a)\theta$ for some $a\in[0,1]$ we can write
\begin{eqnarray*}
D_{T}(\theta'/\lambda,\beta)&\leq& a D_{T}(\beta/\lambda,\beta)+(1-a) D_{T}(\theta/\lambda,\beta)\\
D_{T}(\theta/\lambda,\beta)-D(\theta'/\lambda,\beta)&\geq& a D_{T}(\theta/\lambda,\beta)-a D_{T}(\beta/\lambda,\beta)
\end{eqnarray*}¥
%Note that 
%\begin{eqnarray*}
%\nabla_{\theta} C_{\lambda} (\theta)&=&\nabla_{\theta}\lambda C\left(\frac{\theta}{\lambda}\right)\\
%&=&\nabla_{\theta/\lambda}C\left(\frac{\theta}{\lambda}\right)\\
%&=&\E_{\theta/\lambda}[\phi(x)]
%\end{eqnarray*}¥
\subsection{exponential utility traders with exponential family beliefs}
As before the prediction market is set up with cost function $C_{\lambda}(\theta)$  and payoff for a unit share vector as $\phi(x)$. Thus, the  payoff of a trader who moves the market state from $\theta$ to $\theta'$ is given by $C_{\lambda}(\theta)-C_{\lambda}(\theta')+\delta\phi(x)$ where $\delta=\theta'-\theta$ and the expected utility of the payoff for a trader with exponential belief parameter $\beta$ is given by
\begin{eqnarray*}
&&\E_{x\sim p(x;\beta)}U[C_{\lambda}(\theta)-C_{\lambda}(\theta')+\delta\phi(x)]\\
&=&U[C_{\lambda}(\theta)-C_{\lambda}(\theta+\delta)+\frac{1}{a}(T(\beta)-T(\beta-a\delta))]\\
&=&U[\lambda C(\theta/\lambda)-\lambda C\left(\frac{\theta+\delta}{\lambda}\right)+\frac{1}{a}(T(\beta)-T(\beta-a\delta))]\\
&=&U[\lambda T(\theta/\lambda)-\lambda T\left(\frac{\theta+\delta}{\lambda}\right)+\frac{1}{a}(T(\beta)-T(\beta-a\delta))]
\end{eqnarray*}¥
Here $T$ is the log partition function; for the exponential family prediction market $C=T$. To maximize the utility, we set the gradient of the argument with respect to $\delta$ to 0. 
$$\nabla T\left(\frac{\theta+\delta}{\lambda}\right) = \nabla T(\beta-a\delta)$$
So we have for an optimal choice of $\delta$, 
\begin{eqnarray*}
\frac{\theta+\delta}{\lambda} &=& \beta-a\delta\\
\implies \delta&=&\frac{\beta\lambda-\theta}{1+a\lambda}\\
\implies \theta+\delta&=&\lambda\frac{\beta+a\delta}{1+a\lambda}
\end{eqnarray*}

@@

\newcommand{\ttheta}{\tilde{\theta}}
As we're dealing with LMSR I write $T$ rather than $C$ now where $T$ is the log partition function. The cost function is the liquidity-adjusted version $T_{\lambda}$. Let $\hmu$ be the agent's mean belief and let $\htheta = \grad T^{-1}(\hmu)$ be the corresponding natural parameter. A risk-neutral agent moves the share vector so that the prices match its mean parameter. Therefore, define the \emph{target shares} as $\ttheta = \grad T^{-1}_{\lambda}(\hmu)$. So $\htheta$ and $\ttheta$ are related by
%
$$
\grad T(\htheta) = \grad T_{\lambda}(\ttheta).
$$
%
It's easy to check that $\grad T_{\lambda}(\ttheta) = \grad T(\lambda \ttheta)$ so we have
%
\begin{equation} \label{target}
\ttheta = \htheta / \lambda.
\end{equation}
%
Now using the liquidity-adjusted cost function, Equation~(17) from Theorem~6.2 in the write-up becomes the following, where $\theta$ is the current market state:
%
\begin{equation} \label{update}
\frac{\lambda}{\lambda+a} \ttheta + \frac{a}{\lambda+a} \theta.
\end{equation}
%
As $\lambda$ grows large, the agent moves the market state closer to the \emph{target shares}, rather than its true natural parameter. Note that the target shares themselves depend on $\lambda$, because of~(\ref{target}). This make sense because higher price responsiveness means the agent needs to buy fewer shares to make the market prices match its own expectations. In that sense, it may be conceptually cleaner to write~(\ref{update}) as
%
\begin{equation}
\frac{\lambda}{\lambda+a} \grad T^{-1}_{\lambda}(\hmu) + \frac{a}{\lambda+a} \grad T^{-1}_{\lambda}(\mu).
\end{equation}
%
where $\mu$ are the current market prices.

\end{document}

>>>>

% Metadata Information
\acmVolume{9}
\acmNumber{4}
\acmArticle{39}
\acmYear{2013}
\acmMonth{6}
\usepackage[numbers]{natbib}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{amsthm}
\usepackage{parskip}
\usepackage{booktabs}
%new commands
\newcommand{\la}{\langle}
\newcommand{\ra}{\rangle}
\newcommand{\mX}{\mathcal{X}}
\newcommand{\bR}{\mathbf{R}}
\newcommand{\grad}{\nabla}
\newcommand{\Exp}{\mathbf{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\q}{\mathbf{q}}
\newcommand{\mM}{\ensuremath{\mathcal{M}}}
\newcommand{\hmu}{\ensuremath{\hat{\mu}}}
\newcommand{\mP}{\ensuremath{\mathcal{P}}}
\newcommand{\mY}{\ensuremath{\mathcal{Y}}}
\newcommand{\mS}{\ensuremath{\mathcal{S}}}
\newcommand{\ds}{\displaystyle}
\newcommand{\mF}{\mathcal{F}}
\newcommand{\mD}{\mathcal{D}}
\newcommand{\htheta}{\hat{\theta}}
\newcommand{\ttheta}{\tilde{\theta}}
%---------------------------------------------------------------------------
\newtheorem{thm}{Theorem}[section]
\newtheorem{prop}[thm]{Proposition}
\newtheorem{lem}[thm]{Lemma}
\newtheorem{con}[thm]{Conjecture}
\newtheorem{cor}[thm]{Corollary}
\newtheorem{ex}[thm]{Example}
\newtheorem{qn}[thm]{Question}
\newcommand{\defn}{\bigbreak\noindent{\bf Definition. }}
\newcommand{\undefn}{\bigbreak}
\newcommand{\rmk}{\bigbreak\noindent{\bf Remark. }}
\newcommand{\unrmk}{\bigbreak}
%\newcommand{\proof}{\noindent{\bf Proof. }}
\newcommand{\unproof}{\hfill$\Box$\bigbreak}
\newcommand{\defeq}{\stackrel{\mathrm{def}}{=}}
\newcommand{\holds}[1]{\stackrel{?}{#1}}
\newcommand{\conv}{\mathrm{conv}}
\newcommand{\ie} {{\it i.e. }}
\newcommand{\eg} {{\it e.g. }}
\newcommand{\etal} {{\it et al. }}
\newcommand{\E}{\mathbf{E}}
\newcommand{\betavec}{\pmb{\beta}}
\newcommand{\qvec}{\mathbf{q}}
\newcommand{\lpf}{\mathbf{\psi}} %logpartition function
\newcommand{\family}{\mathcal F} 
\newcommand{\suff}{\pmb{\phi}} % suff stats
\newcommand{\muvec}{\pmb{\mu}}

@@

\section{Cost Function with a Liquidity Parameter}
We may introduce a liquidity parameter to the cost function as $C_{\lambda}(\theta)=\frac{1}{\lambda} C(\lambda\theta)$. We would like to find an expression for a lower bound on expected payoff. We would also like to find an expression for final market state for exponential utility traders with exponential family beliefs.
\subsection{lower bound on expected payoff}
The prediction market is set up with cost function $C_{\lambda}(\theta)$  and payoff for a unit share vector as $\phi(x)$. To find a lower bound on expected payoff, we first note that
\begin{eqnarray*}
\E_{x\sim p(x;\beta)}[C_{\lambda}(\theta)-\theta\phi(x)]&=&C_{\lambda}(\theta)-\theta\nabla T(\beta)\\
&=&\frac{1}{\lambda} T(\theta\lambda)-\theta\nabla T(\beta)\\
&=&\frac{1}{\lambda} [T(\theta\lambda)-\theta\lambda\nabla T(\beta)-T(\beta)+\beta\nabla T(\beta)]+\frac{1}{\lambda} T(\beta)-\frac{\beta}{\lambda}\nabla T(\beta)\\
&=&\frac{1}{\lambda} D_{T}(\theta\lambda,\beta)+\frac{1}{\lambda} [T(\beta)-\beta\nabla T(\beta)]
\end{eqnarray*}¥ 
where $D_{T}(x_{1},x_{2}) = T(x_{1})-T(x_{2})-\nabla T(x_{2})(x_{1}-x_{2})$ is the Bregman Divergence between $x_{1}$ and $x_{2}$ based on $T$.

Thus, the expected payoff  for a trader with exponential family belief parameter $\beta$ who moves the market state from $\theta$ to $\theta'$ can be written as
\begin{eqnarray*}
\E_{x\sim p(x;\beta)}\left[C_{\lambda}(\theta)-\theta\phi(x)-(C_{\lambda}(\theta')-\theta'\phi(x))\right]=\frac{1}{\lambda}[D_{T}(\theta\lambda,\beta)-D_{T}(\theta'\lambda,\beta)]
\end{eqnarray*}¥
We have $\theta'=\frac{\lambda}{\lambda+a}(\beta/\lambda)+\frac{a}{\lambda+a}\theta$ where $a$ is the utility parameter
\begin{eqnarray*}
D_{T}(\theta'\lambda,\beta)&\leq& \frac{\lambda}{\lambda+a} D_{T}(\beta,\beta)+\frac{a}{\lambda+a} D_{T}(\theta\lambda,\beta)\\
D_{T}(\theta/\lambda,\beta)-D_{T}(\theta'/\lambda,\beta)&\geq& \frac{\lambda}{a} D_{T}(\theta'/\lambda,\beta)
\end{eqnarray*}¥

Thus, the expected payoff is at least $\frac{1}{a} D_{T}(\theta'/\lambda,\beta)$.

%Note that 
%\begin{eqnarray*}
%\nabla_{\theta} C_{\lambda} (\theta)&=&\nabla_{\theta}\lambda C\left(\frac{\theta}{\lambda}\right)\\
%&=&\nabla_{\theta/\lambda}C\left(\frac{\theta}{\lambda}\right)\\
%&=&\E_{\theta/\lambda}[\phi(x)]
%\end{eqnarray*}¥
\subsection{exponential utility traders with exponential family beliefs}
As before the prediction market is set up with cost function $C_{\lambda}(\theta)$  and payoff for a unit share vector as $\phi(x)$. Thus, the  payoff of a trader who moves the market state from $\theta$ to $\theta'$ is given by $C_{\lambda}(\theta)-C_{\lambda}(\theta')+\delta\phi(x)$ where $\delta=\theta'-\theta$ and the expected utility of the payoff for a trader with exponential belief parameter $\beta$ is given by
\begin{eqnarray*}
&&\E_{x\sim p(x;\beta)}U[C_{\lambda}(\theta)-C_{\lambda}(\theta')+\delta\phi(x)]\\
&=&U[C_{\lambda}(\theta)-C_{\lambda}(\theta+\delta)+\frac{1}{a}(T(\beta)-T(\beta-a\delta))]\\
&=&U[\lambda C(\theta/\lambda)-\lambda C\left(\frac{\theta+\delta}{\lambda}\right)+\frac{1}{a}(T(\beta)-T(\beta-a\delta))]\\
&=&U[\lambda T(\theta/\lambda)-\lambda T\left(\frac{\theta+\delta}{\lambda}\right)+\frac{1}{a}(T(\beta)-T(\beta-a\delta))]
\end{eqnarray*}¥
Here $T$ is the log partition function; for the exponential family prediction market $C=T$. To maximize the utility, we set the gradient of the argument with respect to $\delta$ to 0. 
$$\nabla T\left(\frac{\theta+\delta}{\lambda}\right) = \nabla T(\beta-a\delta)$$
So we have for an optimal choice of $\delta$, 
\begin{eqnarray*}
\frac{\theta+\delta}{\lambda} &=& \beta-a\delta\\
\implies \delta&=&\frac{\beta\lambda-\theta}{1+a\lambda}\\
\implies \theta+\delta&=&\frac{1}{1+a\lambda}\beta\lambda+\frac{a\lambda}{1+a\lambda}\delta
\end{eqnarray*}

From the remarks below, we see that $\beta\lambda$ actually represents the agent's target share vector.

@@

As we're dealing with LMSR I write $T$ rather than $C$ now where $T$ is the log partition function. The cost function is the liquidity-adjusted version $T_{\lambda}$. Let $\hmu$ be the agent's mean belief and let $\htheta = \grad T^{-1}(\hmu)$ be the corresponding natural parameter. A risk-neutral agent moves the share vector so that the prices match its mean parameter. Therefore, define the \emph{target shares} as $\ttheta = \grad T^{-1}_{\lambda}(\hmu)$. So $\htheta$ and $\ttheta$ are related by
%
$$
\grad T(\htheta) = \grad T_{\lambda}(\ttheta).
$$
%
It's easy to check that $\grad T_{\lambda}(\ttheta) = \grad T(\lambda \ttheta)$ so we have
%
\begin{equation} \label{target}
\ttheta = \htheta / \lambda.
\end{equation}
%
Now using the liquidity-adjusted cost function, Equation~(17) from Theorem~6.2 in the write-up becomes the following, where $\theta$ is the current market state:
%
\begin{equation} \label{update}
\frac{\lambda}{\lambda+a} \ttheta + \frac{a}{\lambda+a} \theta.
\end{equation}
%
As $\lambda$ grows large, the agent moves the market state closer to the \emph{target shares}, rather than its true natural parameter. Note that the target shares themselves depend on $\lambda$, because of~(\ref{target}). This make sense because higher price responsiveness means the agent needs to buy fewer shares to make the market prices match its own expectations. In that sense, it may be conceptually cleaner to write~(\ref{update}) as
%
\begin{equation}
\frac{\lambda}{\lambda+a} \grad T^{-1}_{\lambda}(\hmu) + \frac{a}{\lambda+a} \grad T^{-1}_{\lambda}(\mu).
\end{equation}
%
where $\mu$ are the current market prices.
\end{document}

Thus, the expected payoff  for a trader with exponential family belief parameter $\beta$ who moves the market state from $\theta$ to $\theta'$ can be written as
\begin{eqnarray*}
\E_{x\sim p(x;\beta)}\left[C_{\lambda}(\theta)-\theta\phi(x)-(C_{\lambda}(\theta')-\theta'\phi(x))\right]=\lambda[D_{T}(\theta/\lambda,\beta)-D_{T}(\theta'/\lambda,\beta)]
\end{eqnarray*}¥
If $\theta'=a\beta\lambda+(1-a)\theta$ for some $a\in[0,1]$ we can write
\begin{eqnarray*}
D_{T}(\theta'/\lambda,\beta)&\leq& a D_{T}(\beta\lambda/\lambda,\beta)+(1-a) D_{T}(\theta/\lambda,\beta)\\
D_{T}(\theta/\lambda,\beta)-D_{T}(\theta'/\lambda,\beta)&\geq& a D_{T}(\theta/\lambda,\beta)-a D_{T}(\beta,\beta)
\end{eqnarray*}¥
